Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-18 doc/mc-paper.tex
Polished main part of paper

\emph{Knowledge compilation} is the process of converting a Boolean
formula into a form that makes it possible to calculate associated
values efficiently, such as the weighted or unweighted model
count. Existing knowledge compilers
provide no guarantee that the representation they generate
correctly represents the original formula, and therefore they cannot validate the generated counts.

@@

We have developed a program that generates POG representations from
the d-DNNF graphs generated by the state-of-the-art knowledge compiler
\dfour, as well as checkable proofs that the output POGs
are equivalent to the input CNF formulas.  We have implemented an
efficient checker that can handle large proofs. Our tool chain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs generated by \dfour{} for formulas from a recent model counting
competition. We also present a checker and a model counter for POGs that have
been formally verified in the \lean{} proof assistant.  These programs both prove the soundness of our proof framework and
create the first formally verified toolchain for weighted and unweighted model counting.
\end{abstract}

@@

Modern Boolean satisfiability (SAT) solvers provide the ability to
find a satisfying assignment to a Boolean formula or to prove that no
such assignment exists.  They have applications across a variety of
domains, including formal verification of hardware, software, and
security protocols, computational mathematics, and combinatorial
optimization.  Some problems of interest, however, require going
beyond standard Boolean satisfiability.  For example, the {\em model
  counting problem} requires computing the number of satisfying
assignments of a formula, including in cases where there are far too many
to enumerate individually.  Model counting, in turn, has
applications in artificial intelligence, computer security, and
statistical sampling.  There are also many useful extensions of standard model counting,
including {\em
  weighted model counting}, where a weight is defined for
each possible assignment, and the goal becomes to compute the sum of the weights
of the satisfying assignments.

@@

One shortcoming of existing knowledge compilers is that they have no
way to validate that their generated results are correct, i.e., that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers can generate
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has also
shown that having this capability gives developers the ability quickly
detect and identify bugs in their solvers, and this, in turn, has led
to more reliable SAT solvers.

In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations commonly used in knowledge
compilation, including d-DNNF\@.  We present the CRAT file format.
It captures both the structure of a POG and a checkable proof
that the POG is logically equivalent to a Boolean formula given in
conjunctive normal form (CNF).  A proof consists of a sequence of
clause addition and deletion steps, based on an extended resolution
proof system~\cite{Tseitin:1983}.  We present a set of conditions that
a CRAT file must satisfy to provide an equivalence proof.

@@

Figure~\ref{fig:chain} illustrates the certifying knowledge compilation
and (weighted) model counting tool chain that we have developed.  A
{\em proof generator} generates a CRAT description from a d-DNNF
graph generated by \dfour{}, a state-of-the-art knowledge compiler.
A {\em proof checker} then verifies that the combination of CNF and CRAT files
satisfy our rules, and a {\em ring evaluator} can compute
either a standard or weighted model count from the POG\@ representation.
We evaluate the capabilities of this tool chain using
benchmark formulas from the 2022 standard and weighted model
competitions.  Our tools are able to handle all but the largest d-DNNF
graphs generated by \dfour{}.
As the dashed box in Figure~\ref{fig:chain} indicates, this tool chain
shifts the need for trustworthy code away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will also be caught by the proof checker.

To further enhance the trustworthiness of our toolchain, we have
developed formally verified versions of the proof checker and ring
evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.  Running these two programs
has the effect of formally verifying both the logical equivalence of
the POG to the input formula and the correctness of the count.  In
addition, the existence of a formally verified checker serves the
important role of establishing the soundness of our proof framework.
Indeed, in the course of developing a formally verified checker, we
identified subtle aspects of the proof obligations required to assure
soundness.

@@

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound.  They also
present experimental results for cases where they modify existing
model counters to also generate proof traces, and where they convert
d-DNNF graphs into a checkable form.  Our tool chain is not directly
comparable to theirs---they certify the output of a standard modeling
program, while we certify the output of a knowledge compiler and then
produce a certified standard or weighted model count.  Their tool
chain can be applied to some programs that are pure model counters,
while ours can be applied to pure knowledge compilers.  Nonetheless,
we compare the performance of our tool chain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.  Our
experience with developing a formally verified proof checker has shown
us that, even for the well understood framework of extended
resolution, it can be challenging to formulate the full set of requirements to guarantee soundness.

We know of no other
work addressing the question of certifying the output of a knowledge
compiler.

@@

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{i=1,k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{i=1,k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

@@

  A {\em partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to a product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to a sum must have disjoint models.  That is, operation
        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

We propose a general framework for summarizing properties of Boolean
formulas inspired by the formulation of {\em algebraic model counting}~\cite{kimmig:jal:2017}.

@@

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.

The {\em weighted model counting}  problem is also defined over
rational numbers.  Some formulations  allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can cast this as a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and computing the weighted count
as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}
As is described in Section~\ref{subsection:counting}, we have proved these three equations using \lean{}.

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.
As can be seen, a POG is represents a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.

@@

POGs are inspired by the d-DNNF graphs devised by Darwiche for
representing Boolean functions \cite{darwiche:jair:2002}.
POGs generalize d-DNNF in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF can have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  These generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  Clauses can also be added or deleted based on
{\em  implication redundancy}.  That is, when clause $C$ satisfies $\phi \imply C$ for some formula $\phi$, then it can either be added to form the set $\phi \cup \{C\}$ or it can be deleted from that set to form $\phi$.

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{wetzler14_drattrim,RAT} for propositional logic. It provides a simple and efficient
way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$ having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

@@

A CRAT file provides both a declaration of a POG, as well as a checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG\@.
The proof format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then becomes a way
of representing constant $0$.

>>>>

\emph{Knowledge compilation} is the process of converting a Boolean
formula into a form that makes it possible to calculate associated
values efficiently, such as the weighted or unweighted model
count. Existing knowledge compilers
provide no guarantee that the representation they produce
correctly represents the original formula, and therefore they cannot validate the generated counts.

@@

We have developed a program that generates POG representations from
the d-DNNF graphs produced by the state-of-the-art knowledge compiler
\dfour, as well as checkable proofs that the output POGs
are equivalent to the input CNF formulas.  We have also implemented an
efficient checker for these proofs. Our tool chain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs produced by \dfour{} for formulas from a recent model counting
competition. We also present a checker and a model counter for POGs that have
been formally verified in the \lean{} proof assistant.  These programs both prove the soundness of our proof framework and
create the first formally verified toolchain for weighted and unweighted model counting.
\end{abstract}

@@

Modern Boolean satisfiability (SAT) solvers can
find a satisfying assignment to a Boolean formula or generate a proof that no
such assignment exists.  They have applications across a variety of
domains, including computational mathematics, combinatorial
optimization, and the formal verification of hardware, software, and
security protocols.  Some applications, however, require going
beyond standard Boolean satisfiability.  For example, the {\em model
  counting problem} requires computing the number of satisfying
assignments of a formula, including in cases where there are far too many
to enumerate individually.  Model counting, in turn, has
applications in artificial intelligence, computer security, and
statistical sampling.  There are also many useful extensions of standard model counting,
including {\em
  weighted model counting}, where a weight is defined for
each possible assignment, and the goal becomes to compute the sum of the weights
of the satisfying assignments.

@@

One shortcoming of existing knowledge compilers is that they have no
way to validate that their generated results are correct, i.e., that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers can generate
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has also
shown that having this capability gives SAT solver developers an ability to quickly
detect and diagnose bugs in their programs, and this, in turn, has led
to more reliable SAT solvers.

In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations produce by current knowledge
compilers, including d-DNNF\@.  We present the CRAT file format.
It captures both the structure of a POG and a checkable proof
that the POG is logically equivalent to a Boolean formula given in
conjunctive normal form (CNF).  A proof consists of a sequence of
clause addition and deletion steps, based on an extended resolution
proof system~\cite{Tseitin:1983}.  We present a set of conditions that
a CRAT file must satisfy to provide an equivalence proof.

@@

Figure~\ref{fig:chain} illustrates the certifying knowledge compilation
and (weighted) model counting tool chain that we have developed.  A
{\em proof generator} generates a CRAT description from a d-DNNF
graph produced by \dfour{}, a state-of-the-art knowledge compiler.
A {\em proof checker} then verifies that the combination of CNF and CRAT files
satisfy our rules, and a {\em ring evaluator} can compute
either a standard or weighted model count from the POG\@ representation.
We evaluate the capabilities of this tool chain using
benchmark formulas from the 2022 standard and weighted model
competitions.  Our tools are able to handle all but the largest d-DNNF
graphs generated by \dfour{}.
As the dashed box in Figure~\ref{fig:chain} indicates, this tool chain
shifts the need for trustworthy code away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will also be caught by the proof checker.

To further enhance the trustworthiness of our toolchain, we have
developed formally verified versions of the proof checker and ring
evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.  Running these two programs
has the effect of formally verifying both the logical equivalence of
the POG to the input formula and the correctness of the count.  In
addition, the existence of a formally verified checker serves the
important role of establishing the soundness of our proof framework.
Our
experience with developing a formally verified proof checker has shown
us that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of requirements to guarantee soundness.
Our efforts to formalize the framework
identified subtle details for some of the proof obligations.

@@

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound.  They also present
experimental results for cases where they modify existing model
counters to generate proof traces, and where they generate proof
traces from d-DNNF graphs.
%% They have not formally verified their
%%proof framework or created a formally verified checker.

Our tool chain is not directly
comparable to theirs---they certify the output of a standard modeling
program, while we certify the output of a knowledge compiler and then
produce a certified standard or weighted model count.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our tool chain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

We know of no other
work addressing the question of certifying the output of a knowledge
compiler or model counter.

@@

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{i=1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{i=1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{i=1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{i=1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

@@

  A {\em partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{i=1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

We propose a general framework for summarizing properties of Boolean
formulas along the lines of algebraic model counting~\cite{kimmig:jal:2017}.

@@

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
weight function $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.

The {\em weighted model counting}  problem is also defined over
rational numbers.  Some formulations  allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can cast this as a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with weight function $w(x) = W(x)/r(x)$ for each
variable $x$, and computing the weighted count
as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{i=1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{i=1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1 \leq i \leq k} \phi_i, w \right) &=& \prod_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1 \leq i \leq k} \phi_i, w\right) &=& \sum_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}
As is described in Section~\ref{subsection:counting}, we have proved these three equations using \lean{}.

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.

As can be seen, a POG is represents a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.

@@

POGs generalize the d-DNNF graphs in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  These generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  Clauses can also be added or
deleted based on {\em implication redundancy}.  That is, when clause
$C$ satisfies $\phi \imply C$ for formula $\phi$, then it can either
be added $\phi$ to create the formula $\phi \cup \{C\}$ or it can be deleted
from $\phi \cup \{C\}$ to create $\phi$.

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$ having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

@@

A CRAT file provides both a declaration of a POG, as well as a checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG\@.
The proof format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
  encoding the product or sum operation.
%  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness
%  Generalizing to $k$ arguments would require
%  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then provides a way
to representing constant $0$.
